%%
%% Author: DKron
%% 24.07.2018
%%

\chapter{Lower Bounds}\label{ch:lowerBounds}

In this chapter we want to establish that the lower bounds of the expressive power of PHFL$^k$ and PHFL$^k_{tail}$ are \exptime{$k$} and \expspace{$k$}, respectively. The
lower bounds of PHFL$^k$ and PHFL$^k_{tail}$ can be proven by encoding the run of a Turing Machine as query.
As another possibility one can use intermediate logics HO(LFP)$^{k+1}$ and HO(PFP)$^{k+1}$ and encode the bisimulation-invariant fragments of these as PHFL$^k$ and
PHFL$^k_{tail}$, respectively. Note that PHFL cannot distinguish between bisimilar structures~\cite{viswanathan2004higher}. This means that PHFL formulas can only define bisimulation-invariant
graph problems. Moreover, PHFL is sufficient to encode the bisimulation-invariant fragments of HO(LFP)$^{k+1}$ and HO(PFP)$^{k+1}$. To encode the bisimulation-invariant fragments of HO(LF\-P)$^{k+1}$ and HO(PFP)$^{k+1}$ as PHFL$^k$ and
PHFL$^{k}_{tail}$ respectively we want to translate HO(LFP)$^{k+1}$ formulas into PHFL$^k$ formulas and 
HO(PFP)$^{k+1}$ into a PHFL$^k_{tail}$ formula. 

Encoding existential quantifiers is the most complex part of the translation. In the 
first section we consider some preparations that are necessary to model them in PHFL. Thereafter,  we show that existential 
quantifiers that bind a variable of order $k \geq 1$ can be expressed by a PHFL$^{k-1}$ formula. In the subsequent section we 
use this formula to show that the bisimulation invariant fragment of HO(LFP)$^{k+1}$ can be encoded into PHFL$^k$ and so 
that the lower bound of the expressive power of PHFL$^k$ is \exptime{$k$}. And finally we show that the lower bound 
of the expressive power of PHFL$^k_{tail}$ is \expspace{$k$}.

\input{lower_bounds/lower_bounds_preparation}

\input{lower_bounds/lower_bounds_existential_quantifiers_in_phfl}

\input{lower_bounds/lower_bounds_phfl_k}

\input{lower_bounds/lower_bounds_phfl_k_plus_1_tail}
